A Survey of Physical Design Techniques of Information Systems

### 2010-07-09 trunk/cs848-db/Paper.tex
finished intro for dw

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Lots of
research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is because the big difference between
processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases. Accordingly, we will see that many of
the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In summary, a main memory index should be
able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also be noted that unlike disk based systems
where it is advantageous to store actual attribute values in the index, main memory systems do not require that. Instead, pointers to the actual attribute
values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

@@

\subsubsection*{B Trees}
Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
performance.

\subsubsection*{T Trees}
Lehmen and Carey introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to main memory.
Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the performance of the T
Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics.

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{figs/Ttree.png}
\caption{T Tree by Lehmen and Carey~\cite{lehman1986study}}
\label{fig:ttree}
\end{figure}

\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}
\cite{rao1999cache}.

\subsubsection*{Cache Sensitive T-Trees (CST-Trees)}

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
with the cache line size. The other problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
Additionally, record pointers stored in the tree take up a lot of space.

@@

\subsubsection*{Cache Sensitive B+ trees (CSB+ Trees)}

Rao and Ross~\cite{rao1999cache} show that $B^{+}-trees$ are more cache conscious than binary search trees and T-Trees which means they are more suited to main
memory databases. However, traditional B+-Trees still had poor cache performance. Therefore, they propose Cache Sensitive B+-Trees
($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.

\subsubsection*{$Prefetching B^{+}-Trees (pB^{+}-trees)$}

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of $B^{+}-trees$. Since current computer systems can prefetch
different data simultaneously, the author take advantage of this. The node size of the $B^{+}-tree$ is made wider than the cache line size. This way, several
cache misses can be overlapped at the same time thus improving performance.

\subsubsection*{$J^{+}-trees$ and $pJ^{+}-trees$}

@@

Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data and load tens of terabytes of
data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring
the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such environment.
Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and
aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further. Moreover, the results
have to be delivered interactively to the business analyst using the system.

The data warehouse supports on-line analytical processing (OLAP), the functional and performance requirements of which are quite different from those of the
on-line transaction processing (OLTP) applications traditionally supported by the operational databases. Given the above challenges, the data warehouse is
usually maintained separately from the organization’s operational databases \cite{248616}. Query throughput and response times are more important than
transaction throughput.

OLAP vs OLTP

OLTP \cite{sen2005comparison}

Data warehouses, in contrast, are targeted for decision support. Historical, summarized and consolidated data is more important than detailed, individual records.

The workloads are query intensive with mostly ad hoc, complex queries that can access millions of records and perform a lot of scans, joins, and aggregates.
Query throughput and response times are more important than transaction throughput.

Operational databases are finely tuned to support known OLTP workloads.

Understanding trends or making predictions requires historical data, whereas operational databases store only current data.

Decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats.

Supporting the multidimensional data models and operations typical of OLAP requires special data organization, access methods, and implementation methods, not
generally provided by DBMSs targeted for OLTP.

@@

- joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}

>>>>

Two considerations are taken when designing index structures for MMDB. The first is that the data resides in main memory and not on disk, and so many of the
considerations taken for I/O operations is no longer there. Instead of focusing on disk access and disk storage, the data structures for main memory
databases should focus on the efficient use of CPU cycles and memory~\cite{lehman1986study}. The second is having index structures cache conscious. Cache
memories improve performance by holding recently referenced data~\cite{smith1982cache}. If the memory reference is found in the cache, then execution proceeds
at processor speed. Otherwise, the data has to be fetched from main memory. With the big difference between processor speed and main memory speed, cache misses
are still relatively expensive. Lots of research work has shown that cache performance is very critical in MMDBs~\cite{boncz1999database,rao1999cache}. This is
because the big difference between processor speed and main memory access speed. Thus, a cache miss is still relatively expensive in main memory databases.
Accordingly, we will see that many of the index data structures proposed for main memory databases focused on being cache conscious or cache sensitive. In
summary, a main memory index should be able to reduce overall computation time while using as little memory as possible~\cite{lehman1986study}. It should also
be noted that unlike disk based systems where it is advantageous to store actual attribute values in the index, main memory systems do not require that.
Instead, pointers to the actual attribute values can be used. This, in turn, will reduce the size of the index which is preferable in this case.

@@

\subsubsection*{B Trees}
Although B Trees, explained in the previous section, are originally designed for disk based database systems, they were found to also be useful in
MMDB~\cite{lehman1986study}. In disk based systems, the B+ Tree is more used than the regular B Tree. However, in main memory databases, the B Tree would be
more suitable since there is no gain from keeping all the data in the leaves. In a main memory system, this would only waste space without improving
performance. In a B or B+ Tree, multiple keys are used to search within a node. This means that if the node fits in a cache line, this cache load can satisfy
more than more comparison leading to better cache utilization~\cite{rao1999cache}. Therefore, B Trees in general have good cache behavior.

However, the cache behavior of B+ Trees could still be improved. Therefore, Rao and Ross~\cite{rao2000making} propose Cache Sensitive B+-Trees
($CSB^{+}-Trees$)~\cite{rao2000making} as an index structure for main memory. To do so, they eliminated most of the child pointers and had more keys in each
node to improve locality and reduce tree height~\cite{luan2009prefetching}. Since the number of cache misses in search operations is proportional to the height
of the tree, $CSB^{+}-trees$ have fewer cache misses, and thus better performance.

Chen et al.~\cite{chen2001improving} propose using prefetching to improve the performance of B+ Trees. Since current computer systems can prefetch
different data simultaneously, the authors take advantage of this. The idea here is to increase the node size of B+ Trees and employ prefectching at the same
time. In a B+ Tree, every time we move down a level in the tree, a cache miss occurs. Accordingly, prefetching larger sizes nodes means that we can overlap
multiple cache misses before they occur. This is because in one cache mix, extra data is retrieved in parallel with the originally requested data if it can be
predicted sufficiently early. Figure~\ref{fig:prefetching} shows how prefetching retrieves the same data in less time. This method, however, does not
eliminate the full cache miss latency with every level in the tree.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{figs/prefetching.png}
\caption{Performance of various B+ Tree searches where a cache miss to memory takes 150 cycles and a subsequent access can begin 10 cycles
later~\cite{chen2001improving}}
\label{fig:prefetching}
\end{figure}

\subsubsection*{T Trees}
Lehmen and Carey~\cite{lehman1986study} introduce a new index structure, the T Tree, which combines the features from AVL Trees and B Trees that are suited to
main memory. Figure~\ref{fig:ttree} shows the proposed structure of the T Tree. They run some experiments on data residing in main memory to compare the
performance of the T
Tree to existing structures. All structures were modified to contain pointers to data values rather than the data values themselves. The structures
compared included  AVL Trees, simple arrays, B Trees, Chained Bucket Hashing, Extendible Hashing, Linear Hashing and Modified Linear Hashing. Their
experiments showed that for unordered data, Modified Linear Hashing gave the best performance, and for ordered data, T Trees gave the best performance for a mix
of searches, inserts and deletes. This is mainly because a T node contains many elements which results in good update and storage characteristics. The T Tree
was the first index structure specifically designed for main memory databases.

Although T Trees have been specifically designed for MMDB, they did not consider the cache behavior at that time~\cite{rao1999cache}. Although many keys are
fit into one node, only the two end keys are used for comparison leading to low node utilization. Similarly, array binary search trees have the same problem.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{figs/Ttree.png}
\caption{T Tree by Lehmen and Carey~\cite{lehman1986study}}
\label{fig:ttree}
\end{figure}

T Trees were the main index structure used for main memory databases for some time after their proposal. However, it was discovered that B+-trees outperform
T-Trees on modern processors because of the growth of cache miss latency relative to processor speed~\cite{rao1999cache,lee2007cst}. This is because the height
of the tree is high which makes the total number of memory accesses from the root to the leaf node higher. The other reason is that the node size is not aligned
with the cache line size. Another problem with T-Trees is that there is a big waste of space in the cache because unnecessary data is brought into the cache.
Additionally, record pointers stored in the tree take up a lot of space.

@@

\subsubsection*{Cache Sensitive Search Trees (CSS-Trees)}

Binary Search Trees have poor cache behavior when the array is much bigger than the cache. Rao and Ross~\cite{rao1999cache} propose the Cache Sensitive Search
Tree (CSS-Tree) to improve this cache behavior. To do that, the search tree is stored as a directory in an array. The number of keys per node is chosen such
that the whole node fits in a cache line. This improves local searching within a node to just one cache miss. They also hard-code the traversal within a node
such that finding the next node happens in $log_{m+1}n$ times rather than $log_{2}n$ times in a binary search tree. However, CSS Trees are only suitable for
read environments since they eliminate nearly all child pointers, thus removing the support for incremental update.

\subsubsection*{J+-trees and pJ+-trees}
\cite{luan2009prefetching}

@@

Data warehouses tend to be extremely large, in fact it is quite possible for a data warehouse to store tens of petabytes of data while loading tens of terabytes of data everyday \cite{thusoo2010data}. Datta et al. \cite {628286} notes that the information in a data warehouse is usually multidimensional in nature, requiring the capability to view the data from a variety of perspectives. Aggregated and summarized data become more crucial than detailed records in such environment. Therefore, the workloads are query intensive with mostly ad hoc, complex queries, often requiring computationally expensive operations such as scans, joins, and aggregation. Performing such operations on large amounts of data, like in the case of data warehousing, complicates the situation further. Moreover, the results have to be delivered interactively to the business analyst using the system.

Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the organization's operational databases \cite{sen2005comparison, 248616}. This architecture can be justified owing to the fact that operational databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP) which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{248616}. Finally, in data warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{figs/dw.png}
\caption{Data Warehousing Architecture \cite{248616}}
\label{fig:dw}
\end{figure}

@@

Joins in data warehouses are usually expensive since the fact table (the largest one) participates in every join \cite{628286}
